Highly Available Transactions: Virtues and Limitations

### 2013-03-31 writeup/hat-analysis-af.tex
PDB pass plus Alan comments

\section{Highly Available Transactions}
\label{sec:hats}

HAT systems, or systems providing transactions with transactional
availability with high availability ($1$-available) or sticky high
availability ($1$-sticky-available), offer substantial latency and
availability benefits, yet they come with a cost to achievable
semantics. In this section, we delineate which of several ACID,
distributed consistency, and session consistency levels can be
achieved with high availability (transactional atomicity, variants of
Repeatable Read isolation, and many session guarantees) and sticky
high availability (read your writes, PRAM and causal consistency) and
which cannot (preventing Lost Update and Write Skew, recency).  We
present a full summary of these results in
Section~\ref{sec:hat-summary}.

As Brewer states, ``systems and database communities are separate but
overlapping (with distinct vocabulary)''~\cite{brewer-slides}. With
this challenge in mind, we draw on existing properties and definitions
from the database and distributed systems literature, providing a
brief, informal explanation and example for each guarantee. In
particular, for our ACID guarantees, we draw largely on Atul Adya's
dissertation~\cite{adya}, Berenson et al.'s 1995 critique of the ANSI
SQL Specification~\cite{ansicritique}, and the ANSI SQL
specification~\cite{ansi-sql}. We provide a set of formal definitions
and semantics in our extended Technical Report~\cite{hat-tr}.

\subsection{Achievable HAT Semantics}

To begin, we present achievable semantics, offering proof-of-concept
algorithms to demonstrate feasibility. These guarantees are fairly
straightforward and have been introduced---albeit with greater brevity
and discussion omitted---in a preliminary workshop
paper~\cite{hat-hotos}. In our examples, we exclusively consider read
and write operations. We denote a write of value $v$ to data item $d$
as $w_d(v)$ and a read from data item $d$ returning $v$ as $r_d(v)$
and assume that all data items have the null value, $\bot$, at
database initialization. Unless otherwise specified, all example
transactions commit.

\subsubsection{ACID Isolation Guarantees}

To begin, Adya's \textit{Read Uncommitted} isolation requires a total
order among transactions and requires that each transaction's writes
are ordered consistently with this order (prohibiting ``Dirty
Writes,'' or $G0$)~\cite{adya}. If two transactions write to the same
set of data items, then the final database state cannot contain the
``earlier'' transaction's writes to the data items. For example, in
the below example, $T_3$ should eventually only read $a=b=1$ or
$a=b=2$ but not $a=2, b=1$ or $a=1, b=2$:
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)
\\T_2 &: w_x(2)~w_y(2)
\\T_3 &: r_x(a)~r_y(b)\\[-2em]
\end{align*}
We will strengthen this property in later models, which will prevent,
for instance, reading $a=2, b=1$ at any time (not just
``eventually''). The difference between this total order on
transactions and the total order required by serializability is that
the Read Uncommitted total order is completely arbitrary (i.e., an
order must exist), whereas a serializable total order must be
equivalent to a serial execution. Read Uncommitted is easily achieved
via applying the same logical timestamp to each update in a
transaction and applying a ``last writer wins'' conflict
reconciliation policy at each replica.

\textit{Read Committed} isolation requires that transactions do not
read uncommitted versions of data items (prohibiting both ``Dirty
Writes''---as above---and ``Dirty Reads'' phenomena; captured by
Adya's $G1a-c$, ANSI's $P1$, and ``broad'' $P1$ (2.2) from Berenson et
al.). For instance, in the example below, $T_3$ should never see
$a=1$, and, if $T_2$ aborts, $T_3$ should never read $a=3$:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_x(2)
\\T_2 &: w_x(3)\\
T_3 &: r_x(a)\\[-2em]
\end{align*}
It is fairly easy to prvent ``Dirty Reads'': if each transaction never
writes uncommitted data to the database, then transactions will never
read each others' dirty data. As a simple solution, clients can buffer
their writes until they commit, or, alternatively, can send them to
servers, who will not serve new writes until clients notify them that
the writes have been committed.

\textit{Repeatable Read} isolation is a contentious property. As
Berenson et al. discuss~\cite{ansicritique}, Gray's original
Repeatable Read lock-based implementation provides substantially
richer guarantees than those that are required by the ANSI SQL
specification~\cite{gray-isolation}. Gray~\cite{gray-isolation},
Berenson et al.~\cite{ansicritique}, and Adya~\cite{adya} all
interpret Repeatable Read as providing serializability for all
operations except predicate-based reads. However, in its description
of Repeatable Read, the ANSI SQL specification provides a useful
property which, although is not true to Gray's spirit of Repeatable
Read, is also commonly found in distributed consistency models: the
ANSI Repeatable Read guarantee requires that, along with observing
Read Committed isolation, each transaction will only read one version
of each data item that it did not itself produce (preventing ``Fuzzy
Read,'' or P2). In the example below, $T_3$ must read $a=1$:
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: w_x(2)
\\T_3 &: r_x(1)~r_x(a)
\end{align*}
This isolation property is a literal interpretation of the phrase
``Repeatable Read'': unless a transaction modifies a given data item,
the observable value of the data item should not change during the
transaction. By itself, this property is rather weak---we can always
read $\bot$ for each data item. However, when coupled with additional
properties, like transactional atomicity, it is stronger: the repeated
reads must obey additional ordering constraints
(Section~\ref{sec:discussion}). In fact, distributed systems often
describe a ``consistent snapshot'' across a set of related events as a
\textit{cut} across a set of participants or, in our case, data
items. Accordingly, to capture the notion that the transaction should
read from a (non-changing) consistent cut over data items (and to
disambiguate from the aforementioned stronger Repeatable Read
properties), we call ANSI SQL ``Repeatable Read'' (i.e., preventing
P2) \textbf{Cut Isolation}. It is possible to satisfy cut isolation
with high availability by caching appropriate versions for each of a
transaction's reads or, alternatively, using multi-versioning on each
replica and ensuring that each of a transaction's successive reads
return the same version of each data item.

We can further consider two variants of Cut Isolation. The first is
relatively straightforward and provides a cut across individual data
items. We will call this \textit{Item Cut Isolation}. The second
provides a cut over individual data items and over logical ranges of
data items, or predicate-based reads (e.g., \texttt{SELECT
  WHERE}). This prevents the \textit{Phantom Problem}, whereby two
successive predicate-based reads return different
data~\cite{gray-isolation}. We call predicate-based cut isolation,
which prevents Phantoms, \textit{Predicate Cut
  Isolation}.\footnote{Oracle provides an isolation level called
  ``Statement Level Read Consistency''; this is analogous to Predicate
  Cut Isolation at the level of a single operation within a
  multi-operation transaction. We do not consider this isolation model
  here as it is subsumed by standard Snapshot Isolation and we believe
  our discussion is easily extended to incorporate it.}

\subsubsection{ACID Atomicity Guarantees}

Transactional atomicity (TA) is core to ACID guarantees. Although, at
least by the ACID acronym, it is not an ``isolation'' property,
transactional atomicity restricts transactions' ability to view the
effects of partially completed transactions. Under transactional
atomicity, within each transaction, either all effects of another
transaction are observed, or none are (equivalently, once some of the
effects of a transaction are observed, all effects are observed). This
is a strictly stronger guarantee than Read Committed isolation: in
Read Committed, we can read a subset of another transaction's
committed writes whereas, with TA, we must read all or none of
them. Together with item cut isolation, TA prevents Read Skew
anomalies (Berenson et al.'s A5A~\cite{ansicritique}). As an example
of TA, $T_2$ must observe $b=c=1$ (or later versions for each key):
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)~w_y(1)~w_z(1)
\\T_2 &: r_x(a)~r_y(1)~r_x(b)~r_z(c)~\\[-1.5em]
\end{align*}
$T_2$ can also observe $a=\bot$, $a=1$, or a later version of
$a$. Notably, TA requires Read Committed isolation: observing all
effects of a transaction implicitly requires observing the final
(committed) effects of a transaction as well.

Perhaps perplexingly, discussions of TA are absent from existing
discussions of weak isolation. This is perhaps again due to the
single-node context in which prior work was developed: on a single
server (or a fully replicated database), TA is achievable via
lightweight locking and/or local concurrency control over data
items. In contrast, in a distributed environment, TA over arbitrary
groups of non-colocated items is considerably more difficult to
achieve with high availability: servers need to know when all of a
transaction's updates are present on their respective
replicas. However, replicas do not need to agree on when to reveal a
new value to clients, which would require consensus; accordingly, TA
only requires reliable broadcast (with some additional metadata for
clients) and is achievable in a HAT system. We omit a full discussion
of the algorithm and instead refer the reader to the extended version
of this paper~\cite{hat-tr}.

\subsubsection{Session Guarantees}

Our models have not yet considered interactions \textit{across}
transactions: we have not guaranteed any ordering or continuity
between transactions (other than that there exists some arbitrary
ordering). This is a natural concern in a distributed context, and, in
the distributed systems literature, many useful \textit{safety}
guarantees span multiple application actions. In particular,
\textit{session guarantees} are used to describe guarantees across
transactions within a given \textit{session}, ``an abstraction for the
sequence of...operations performed during the execution of an
application''~\cite{sessionguarantees}. Informally a session describes
a context that should persist between transactions: for example, on a
social networking site, all of a user's transactions submitted between
``log in'' and ``log out'' operations might form a session.

There are several session guarantees that we can make with high availability:

\vspace{.5em}\noindent\textit{{Monotonic reads}} requires that,
within a session, subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

\vspace{.5em}\noindent\textit{{Monotonic writes}} requires that
each session's writes be serialized. Any order on transactions should
also respect the order in which each session submitted transactions.

\vspace{.5em}\noindent\textit{{Writes Follow Reads}} requires that, if
a session observes an effect of transaction $T_2$ and subsequently
commits transaction $T_2$, then another session can only observe
effects of $T_2$ if it can also observe $T_1$'s (or later values for
$T_1$'s effects). Any order on transactions should respect the
reads-from order.\vspace{.5em}

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until each write's respective dependencies are
fulfilled on all replicas. This mechanism effectively ensures that all
clients read from a globally agreed upon lower bound on the versions
written to the system. This \textit{is} highly available as a client
will never block due to inability to find a server with a sufficiently
up-to-date version of a data item. However, it does not imply that
transactions will read their own writes or, in the presence of
partitions, make forward progress through the version history. The
problem is that, if a server becomes partitioned, under the highly
available model, we must handle the possibility that an unfortunate
client will be forced to issue her next requests against the
partitioned server!

The solution to this conundrum is to give up high availability in
favor of sticky availability. Sticky availability permits three
additional models, which we first define and then prove are
unachievable in a generic highly available system:

\vspace{.5em}\noindent\textit{{Read your writes}} requires
that whenever a client reads a given data item after updating it, the
read returns the updated value (or a value with a higher ID).

\vspace{.5em}\noindent\textit{{PRAM}} (Pipelined Random Access
Memory) provides the illusion of serializing each session's operations
and is the combination of monotonic reads, monotonic writes, and read
your writes~\cite{herlihy-art}.

\vspace{.5em}\noindent\textit{{Causal
    consistency}}~\cite{causalmemory} is the combination of all of the
session guarantees~\cite{sessiontocausal} (alternatively, PRAM with
writes-follow-reads) and is also referred to as PL-2L
isolation~\cite{adya}).\vspace{.5em}

Read your writes is not achievable in a highly available
($1$-available) system. Consider a client that executes the following
two transactions in succession:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: r_x(a)
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client executes $T_2$ against the same (partitioned)
server, then it will be able to read its writes. However, if the
network topology shifts and the client can only contact a different
server which is partitioned from the server that executed $T_1$, then
the client will be unable to read its own writes and the system will
have to either stall indefinitely to allow the client to read her
writes (violating transactional availability) or will have to
sacrifice read your writes guarantees. However, if the client remains
sticky with the server that executed $T_1$, then we can disallow this
scenario. Accordingly, read your writes, and, by proxy, causal
consistency and PRAM require stickiness. Read your writes is ``free''
in a sticky system, while the remaining causality and PRAM guarantees
can be accomplished with the algorithms for achieving the remaining
session guarantees.

@@

\noindent{\textbf{Consistency}} A HAT system can make limited
consistency guarantees. It can often execute commutative and logically
monotonic~\cite{calm} operations without the risk of invalidating
(also monotonic) application-level integrity constraints. Our goal in
this paper is not to sketch the entire space of consistency models
that are achievable (see Section~\ref{sec:futurework}. We specifically
evaluate TPC-C transaction semantics under HAT consistency guarantees
in Section~\ref{sec:evaluation}.

@@

In this section, we demonstrate that preventing Lost Update and Write
Skew---and therefore providing Snapshot Isolation, Repeatable Read,
and one-copy serializability---requires unavailability.

@@

\textbf{Write Skew} is a generalization of Lost Update to multiple
keys. It occurs when one transaction $T1$ reads a given data item $x$,
a second transaction $T2$ reads a different data item $y$, then $T1$
writes to $y$ and commits and $T2$ writes to $x$ and commits. As an
example of Write Skew, consider the following two transactions:
\begin{align*}
\small
T_1 &: r_y(0)~w_x(1)
\\T_2 &: r_x(0)~w_y(1)
\end{align*}
As Berenson et al. describe, if there was an integrity constraint
between $x$ and $y$ such that only one of $x$ or $y$ should have value
$1$ at any given time, then this write skew would lead to a
non-serializable execution. Write skew in particular is a somewhat
esoteric anomaly---for example, it does not appear in
TPC-C~\cite{snapshot-serializable}---but, as a generalization of Lost
Update, it is also unavailable to HATs.

The inability to prevent Lost Update means that Consistent Read,
Snapshot Isolation, and Cursor Stability guarantees are all
unavailable. The inability to prevent Lost Update or Write Skew means
that Repeatable Read and One-Copy Serializability guarantees are
unavailable~\cite{adya,ansicritique}.

@@

Data storage systems make various recency guarantees on reads of data
items. As we have discussed, one of the most famous is
linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability}~\cite{adya} (e.g., Google's
Spanner~\cite{spanner}). It is also common, particularly in systems
that allow reading from masters and slaves, to provide a guarantee
such as ``read a version that is no more than five seconds out of
date'' or similar. Unfortunately, an indefinitely long partition can
force a system to violate any recency bound, so recency bounds are not
enforceable in HAT systems.

@@

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Interestingly, strong one-copy serializability subsumes all
other models, while considering the (large) power set of all
compatible models (e.g., the diagram depicts 96 possible highly
available combinations) hints at the vast expanse of consistency
models found in the literature. This taxonimization is \textit{not}
exhaustive, but we believe it lends substantial clarity into the
relationships between a large subset of the prominent ACID and
distributed consistency models. Additional read/write transaction
semantics that we have omitted should be easily classifiable based on
the available primitives and HAT-incompatible anomaly prevention we
have already discussed.

>>>>

Read your writes is not achievable in a highly available
($1$-available) system. Consider a client that executes the following
two transactions in succession:
\vspace{-.5em}
\begin{align*}
\small
T_1 &: w_x(1)
\\T_2 &: r_x(a)
\end{align*}
If the client executes $T_1$ against a server that is partitioned from
the rest of the other servers, the server must allow $T_1$ to
commit. If the client executes $T_2$ against the same (partitioned)
server, then it will be able to read its writes. However, if the
network topology shifts and the client can only contact a different
server that is partitioned from the server that executed $T_1$, then
the client will be unable to read its own writes and the system will
have to either stall indefinitely to allow the client to read her
writes (violating transactional availability) or will have to
sacrifice read your writes guarantees. However, if the client remains
sticky with the server that executed $T_1$, then we can disallow this
scenario. Accordingly, read your writes, and, by proxy, causal
consistency and PRAM require stickiness. Read your writes is ``free''
in a sticky system, while the remaining causality and PRAM guarantees
can be accomplished with the algorithms for achieving the remaining
session guarantees.

@@

\noindent{\textbf{Consistency}} A HAT system can make limited
consistency guarantees. It can often execute commutative and logically
monotonic~\cite{calm} operations without the risk of invalidating
(also monotonic) application-level integrity constraints. Our goal in
this paper is not to sketch the entire space of consistency models
that are achievable (see Section~\ref{sec:futurework}). We
specifically evaluate TPC-C transaction semantics under HAT
consistency guarantees in Section~\ref{sec:evaluation}.

@@

In this section, we demonstrate that preventing Lost Update and Write
Skew---and therefore providing Snapshot Isolation, Repeatable Read,
and one-copy serializability---inherently requires unavailability.

@@

\textbf{Write Skew} is a generalization of Lost Update to multiple
keys. It occurs when one transaction $T1$ reads a given data item $x$,
a second transaction $T2$ reads a different data item $y$, then $T1$
writes to $y$ and commits and $T2$ writes to $x$ and commits. As an
example of Write Skew, consider the following two transactions:
\begin{align*}
\small
T_1 &: r_y(0)~w_x(1)
\\T_2 &: r_x(0)~w_y(1)
\end{align*}
As Berenson et al. describe, if there was an integrity constraint
between $x$ and $y$ such that only one of $x$ or $y$ should have value
$1$ at any given time, then this write skew would violate the constraint (which is preserved in serializable executions). Write skew is a somewhat
esoteric anomaly---for example, it does not appear in
TPC-C~\cite{snapshot-serializable}---but, as a generalization of Lost
Update, it is also unavailable to HATs.

Their need to prevent Lost Update means that Consistent Read, Snapshot
Isolation, and Cursor Stability guarantees are all unavailable.
Repeatable Read (in the sense of Gray~\cite{gray-isolation}, Berenson
et al.~\cite{ansicritique}, or Adya~\cite{adya}) and One-Copy
Serializability need to prevent both Lost Update and Write Skew; this
means that they are also inherently unavailable.

@@

Data storage systems make various recency guarantees on reads of data
items. As we have discussed, one of the most famous is
linearizability~\cite{herlihy-art}, which states that reads will
return the last completed write to a data item, and there are several
other (weaker) variants such as safe and regular register
semantics. When applied to transactional semantics, the combination of
one-copy serializability and linearizability is called \textit{strong
  (or strict) one-copy serializability}~\cite{adya} (e.g., Google's
Spanner~\cite{spanner}). It is also common, particularly in systems
that allow reading from masters and slaves, to provide a guarantee
such as ``read a version that is no more than five seconds out of
date'' or similar. Unfortunately, an indefinitely long partition can
force an available system to violate any recency bound, so recency
bounds are not enforceable in HAT systems.

@@

To the best of our knowledge, this is the first unification of
database ACID, distributed consistency, and session guarantee
models. Interestingly, strong one-copy serializability subsumes all
other models, while considering the (large) power set of all
compatible models (e.g., the diagram depicts 96 possible highly
available combinations) hints at the vast expanse of consistency
models found in the literature. This taxonomy  is \textit{not}
exhaustive, but we believe it lends substantial clarity into the
relationships between a large subset of the prominent ACID and
distributed consistency models. Additional read/write transaction
semantics that we have omitted should be easily classifiable based on
the available primitives and HAT-incompatible anomaly prevention we
have already discussed.

### 2013-03-31 writeup/model.tex
PDB pass plus Alan comments

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response; this means servers will not need to communicate with any
others. If servers are partitioned from one another, they do not need
to stall in order to provide a ``safe'' query response. This lack of
fast-path coordination also means that a highly available system also
provides low latency~\cite{abadi-pacelc}; in a wide-area setting, this
means that clients need not need to wait for cross-datacenter
communication. However, to properly describe whether a
\textit{transactional} system is highly available, we need both a way
to describe what servers a client must contact as well as what kinds
of responses a server can provide, especially given the possibility of
aborts.

@@

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between
  a \textit{fully replicated} system, in which all servers are
  replicas for all data items, and a \textit{partially replicated}
  system, in which at least one server acts as a replica for only a
  subset of all data items. For generality, and, for applicability to
  many modern ``sharded'' or ``partitioned'' data storage systems like
  Dynamo, we consider partially replicated systems.} a server that can
handle an operation for a given data item is called a \textit{replica}
for that item. While this standard model is useful, there are several
other cases we should consider.

First, clients may wish to contact multiple servers. For example, to
ensure that the effects of operations will persist in the event of
server failures, clients may wish to wait until operations reach more
than one server before they consider the operations
completed. However, according to our definition of high availability,
server-fault tolerance is unavailable. To allow flexibility in the
number of participating servers required for an operation, we propose
\textit{$K$-availability}: if a client can contact $K$ servers, then
it eventually receives a response from all of them, even in the
presence of indefinitely long network partitions. Compared to a system
providing high availability ($1$-availability), a system with
$2$-availability will require more messaging and possibly higher
latency and may experience more unsuccessful operations in the
presence of network and server failures.  As is commmon in standard
consensus algorithms like Paxos, clients may want to contact a
majority of $N$ servers in a system. This \textit{majority
  availability} (a special case of $K$-availability with $K=\lceil
\frac{N}{2} \rceil$) is ``more available'' than full $N$-availability.

Clients may also wish to contact the same server (or set of servers)
across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously maintaining availability
and stickiness entails fate-sharing between the client and its set of
servers: if a client's sticky servers fail, or, equivalently, if the
client is partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a server that has observed all of its prior operations, it eventually
receives a response, even in the presence of indefinitely long
partitions. As above, it may be useful for a client to achieve
stickiness \textit{and} contact many replicas (i.e., for fault
tolerance), so we generalize sticky availability to \textit{$K$-sticky
  availability}, which provides sticky availability with $K$ distinct
servers. A client may choose to become $1$-sticky available by acting
as a server itself; for example, a client might cache its reads and
writes. However, a single client cannot achieve $K$-sticky
availability with $K>1$ without contacting other replicas. Moreover, ``sticking'' all clients
to a designated set of servers (i.e., master availability) is a
special case of sticky availability and is ``less available'' than the
equivalent non-mastered sticky availability.

@@

We show a hierarchy of replica availability levels in
Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
$K$-availability, but $K$-sticky availability is incomparable with
$(K+1)$-availability. $N$-sticky availability is equivalent to
$N$-availability, while majority availability subsumes $1$-sticky
availability. More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
availability subsumes $K$-sticky availability. We have omitted
discussion of operation-specific availability levels (e.g.,
$N$-availability for writes and $1$-availability for reads in a
write-all, read-one data store), system membership changes, or
heterogeneous replicas (e.g., servers in a local and remote
datacenters) but believe there are several avenues for further
taxnomization.

@@

So far, we have focused on single-object, single-operation
availability. This is standard in distributed systems literature
(e.g., linearizable, regular, and safe register models all concern
single objects~\cite{herlihy-art}), yet the database literature
largely focuses on transactions: groups of multiple operations over
multiple objects. Accordingly, by itself, replica availability is
insufficient to describe availability guarantees for
transactions. Additionally, given the choice of \textit{commit} and
\textit{abort} responses---which signal transaction success or failure
to a client---we must be careful in how we define transactional availability.

>>>>

Informally, highly available algorithms ensure ``always on'' operation
and guaranteed low latency. If users of a highly available system are
able to contact a (set of) server(s) in a system, they are guaranteed
a response; this means servers will not need to communicate with any
others. If servers are partitioned from one another, they do not need
to stall in order to provide a ``safe'' query response. This lack of
fast-path coordination also means that a highly available system also
provides low latency~\cite{abadi-pacelc}; in a wide-area setting, this
means that clients need not need to wait for cross-datacenter
communication. To properly describe whether a \textit{transactional}
system is highly available, we need to describe what servers a client
must contact as well as what kinds of responses a server can provide,
especially given the possibility of aborts.

@@

Traditionally, a system provides high availability if every user that
can contact a server eventually receives a response from that server,
even in the presence of arbitrary, indefinitely long network
partitions between servers~\cite{gilbert-cap}. As in a standard
distributed database, designated servers might handle operations on
different data items;\footnote{There is a further distinction between
  a \textit{fully replicated} system, in which all servers are
  replicas for all data items, and a \textit{partially replicated}
  system, in which at least one server acts as a replica for only a
  subset of all data items. For generality, and, for applicability to
  many modern ``sharded'' or ``partitioned'' data storage
  systems~\cite{dynamo, pnuts, bigtable, spanner}, we consider partially
  replicated systems.} a server that can handle an operation for a
given data item is called a \textit{replica} for that item. While this
standard model is useful, there are several other cases we should
consider.

First, clients may wish to contact multiple servers. For example, to
ensure that the effects of operations will persist in the event of
server failures, clients may wish to wait until operations reach more
than one server before they consider the operations
completed. However, according to our definition of high availability,
server-fault tolerance is unavailable. To allow flexibility in the
number of participating servers required for an operation, we propose
\textit{$K$-availability}: if a client can contact $K$ servers, then
it eventually receives a response from all of them, even in the
presence of indefinitely long network partitions. Compared to a system
providing \textit{high availability} ($1$-availability), a system with
$2$-availability will be \textit{less available}, or require more
messaging and possibly higher latency and may experience more
unsuccessful operations in the presence of network and server
failures.  As is commmon in standard consensus algorithms like Paxos,
clients may want to contact a majority of $N$ servers in a
system. This \textit{majority availability} (a special case of
$K$-availability with $K=\lceil \frac{N}{2} \rceil$) is \textit{more
  available} than full $N$-availability.

Clients may also wish to contact the same server (or set of servers)
across subsequent operations. As we will discuss in
Section~\ref{sec:hats}, clients can ensure continuity between
operations (e.g., reading their prior updates to a data item) by
maintaining affinity or ``stickiness'' with a server or set of
servers~\cite{vogels-defs}. Simultaneously maintaining availability
and stickiness entails fate-sharing between the client and its set of
servers: if a client's sticky servers fail, or, equivalently, if the
client is partitioned from its sticky servers, then it may face
unavailability. We say that a system provides \textit{sticky
  availability} if, whenever a client's operation is executed against
a server that has observed all of its prior operations, it eventually
receives a response, even in the presence of indefinitely long
partitions. As above, it may be useful for a client to achieve
stickiness \textit{and} contact many replicas (i.e., for fault
tolerance), so we generalize sticky availability to \textit{$K$-sticky
  availability}, which provides sticky availability with $K$ distinct
servers. A client may choose to become $1$-sticky available by acting
as a server itself; for example, a client might cache its reads and
writes~\cite{bolton, sessionguarantees, swift}. However, a single client
cannot achieve $K$-sticky availability with $K>1$ without contacting
other replicas. Moreover, ``sticking'' all clients to a designated set
of servers (i.e., master availability) is a special case of sticky
availability and is ``less available'' than the equivalent
non-mastered sticky availability.

@@

We show a hierarchy of replica availability levels in
Figure~\ref{fig:availability-order}. $K$-sticky availability subsumes
$K$-availability, but $K$-sticky availability is incomparable with
$(K+1)$-availability. $N$-sticky availability is equivalent to
$N$-availability, while majority availability subsumes $1$-sticky
availability. More generally, $\lceil \frac{N}{2} \rceil$$+$$K$$-$$1$
availability subsumes $K$-sticky availability. We have omitted
discussion of operation-specific availability levels (e.g.,
$N$-availability for writes and $1$-availability for reads in a
write-all, read-one data store), system membership changes, or
heterogeneous replicas (e.g., servers in a local and remote
datacenters) but believe there are several avenues for further
taxonomization.

@@

So far, we have focused on single-object, single-operation
availability. This is standard in distributed systems literature
(e.g., linearizable, regular, and safe register models all concern
single objects~\cite{herlihy-art}), yet the database literature
largely focuses on transactions: groups of multiple operations over
multiple objects. Accordingly, by itself, replica availability is
insufficient to describe availability guarantees for
transactions. Additionally, given the choice of \textit{commit} and
\textit{abort} responses---which signal transaction success or failure
to a client---we must carefully define transactional availability.
